Solution of Differential Algebraic Equations with Local Constraints on the Latent Space of an Autoencoder

### 2019-11-28 paper/paper.tex
Work on the paper
old edits to the paper

\jnlcitation{\cname{Queiruga, A.}\ctitle{Solution of DAEs on Latent Spaces},\cjournal{I. J. Num. Meth. Eng.}, \cvol{2019;00:1--2}}

\keywords{Class file; \LaTeXe; \emph{Wiley NJD}}

@@

\begin{figure}
  \centering
  \includegraphics[width=3in]{../slides/cstr.png}
  \caption{ \label{fig:cstr} Illustration of the Continuous
    Stirred-Tank    
  Reactor. The material in the reactor is exposed to a reservoir with
  pressure
  $p_\infty$ and temperature $T_\infty$ via a conduit with permeability
  and thermal conductivities $k_p$ and $k_T$, direct forced mass
  injection or withdrawal $r$ via a pump, or heating and cooling $s$
  via a heater/heat-exchanger. Stirring and insulation keeps the
  material at a uniform pressure and temperature. Gravity is
  neglected (though is usually the motivation for putting a feeding inlet at
the top and product outlet at the bottom.)}
\end{figure}

@@

Note that the definition of enthalpy as a per-mass quanitity means the mixture needs to be weighted by the
density: $h_{mix}=\sum{\rho_a h_aS_a}/\sum{rho_aS_a}$.

But, what is a phase boundary? The traditional explanation is shown in
Figure \ref{fig:phaseboundaries}. The use of intensive properties breaks down:\footnote{https://en.wikipedia.org/wiki/Phase\_transition}

\begin{quote}
Phase transitions occur when the thermodynamic free energy of a system
is non-analytic for some choice of thermodynamic variables (cf. phases).
\end{quote}

{\bf Remark:} This interpretation breaks down with the existenceof super-saturated phases in non-stable regions: e.g., it is possible to
heat water beyond it's boiling point in a microwave. We are concerned
with environments where this would not happen (the very rough and
heterogenous) pores of rocks, so we have not thought of how to
integrate this problem.

@@

This work focuses on the EOS of pure water. The most accurate
formulations were developed by the International Association for the
Properties of Water Steam (IAPWS) using a variety of experimental
sources for multiple phases \cite{wagner_iapws_2000,
  water_revised_2007, water_revised_2009}. We utilize the 1997
formulation for water gas, liquid, and supercritical phases, and the
2006 formulation for the Ice Ih phase to build the dataset.\footnote{Our
implementation is publicly available at GITHUB:EOS.} (Their
painstaking work in determining the forms of decent fits is a great
motivation for a 21st century approach.)
We can also assemble the surface of possible $T,p,\rho$ coordinates,
shown in Figure \ref{fig:pTrho}.
\begin{figure}
\centering
\includegraphics[width=4in]{../figures/phase_diagram.png}
\caption{\label{fig:phaseboundaries}Phase boundaries for water on the $T-p$ diagram. The standard
interpretation is that the properties of the material change
drastically when crossing this boundary. Note how it is possible to form a path from a
definitively gas state to a definitively liquid state without crossing
a phase boundary. The supercritical regime has no clear boundaries,
but requires a separate curve fit in practice due to numerical challenges.}
\end{figure}

@@

{\bf Remark:} Many popular forms of EOS do not even have closed form
expressions. The family of cubic equations of state for nonideal
gases, such as the Peng-Robinsion equation \cite{peng_new_1976}, are
expressed as functions for $p$ of $V$ (a proxy for $\rho$) and
$T$. Since $\rho$ is rarely a primary varibale, these expressions must be inverted for
$\rho$ given $p$ by solving the cubic equation. This can be performed
analytically, as is done in TOUGH+, but often polynomial roots can
be best solved numerically. When the equations become more
complicated, such as when there are gas mixtures, the equations of
state must be solved numerically. TOUGH+ employs a Richardson
iteration \cite{moridis_users_2014-2}.
Note how this is a numerical
iteration {\em inside} of another numerical iteration that is trying
to differentiation this equation state. Approximations
abound--the mathematical theory is not straight foward. Passing
automatic differentiation objects through these computations is
difficult in practice. Equations are
often solved faster and more accurately by numerical
approximations--why not completely rewrite the equations?

@@

The typical methodology is to use the empirical relations for density
and enthalpy \(h\) as a function of pressure and temperature and solve
the DAE for \(p\) and \(T\) implicitly:
\begin{align}
\text{Solve for}\, p(t)\, \text{and}\, T(t)\, \text{such that:}\\
\partial_t \rho(p,T) & = \nabla \cdot \mathbf{k}\nabla p + r\\
\partial_t \rho(p,T) h(p,T)-p & = \nabla \cdot \mathbf{k'}\nabla T + s
\end{align}
The complication is that the functions \(\rho(p,T)\) and \(u(p,T)\)
are not well defined functions due to the presence of phase changes that
yield sharp discontinuities in \(p,T\), as shown in Figure 1 for
water.
Each state requires a potentially different set of primary
variables. For a pure water system, this would have to be:
\begin{center}
\begin{tabular}{l|c|c}
  State & $X_1$ & $X_2$ \\
  \hline
  Ice & $p$ & $T$ \\
  Liquid & $p$ & $T$ \\
  Gas & $p$ & $T$ \\
  Supercritical & $\rho$ & $T$ \\
  Ice-Liquid & $S_{ice}$ & $T$ \\
  Ice-Gas & $S_{ice}$ & $T$ \\
  Gas-Liquid & $S_{gas}$ & $T$ \\
  Ice-Gas-Liquid & $S_{ice}$ & $S_{gas}$ 
\end{tabular}
\end{center}
Solving the balance laws with this caveat requires the incorpoeration
of a statemachine. Representing the material with the known
representations thus requires two variables and an additional state
tag. For the complete states of water, the states and their possible
transitions is shown in Figure \ref{fig:statemachine}.
  \begin{figure}
    \centering
    \includegraphics[width=4in]{../slides/phase_4.png}
    \caption{\label{fig:statemachine}States and possible transitions for water.}
\end{figure}
Formulate balance equations as
\begin{equation}
R(X ; phase) = 0
\end{equation}
Solve the differentiable part:
\[\frac{\partial R}{\partial X}\Delta X^{k+1/2} = -R(X^k ; phase^k)\]
and iterate on the nondifferential part:
\[phase^{k+1},X^{k+1} = statemachine(X^{k+1/2} ; phase^k)\]
The phase switching logic has the form of a switch case illustrated in
Algorithm \ref{alg:statemachine}.

@@

An archetypical constraint problem is the pendulum. Posed as dynamics
problem with a Lagrange-multipler-esque centripetal force, it
has three unknowns wtih two second-order equations and a constraint,

\begin{align}
\text{Solve for}\, x(t), \, y(t), \, f(t) \, \text{satisfying:} \nonumber \\
m \ddot{x} & = f x/\sqrt{x^2+y^2} \\
m \ddot{y} & = f y/\sqrt{x^2+y^2} - m g \\
  x^2 + y^2 & = R^2
  \text{given } x,y, \dot{x},\dot{y} \text{ at } t=0. \nonumber
\end{align}
Note that we are considering the nonlinear pendulum.

@@

\begin{figure}
  \center
\includegraphics[width=3in]{../figures/autoencoder_balance_detailed.png}
  \caption{\label{fig:balance}Illustration of the insert of the encoder to form the
  description of the system of equation. The equations
  of motion are generated in a declarative programming methodology,
  where the trained decoder can be extracted and inserted into the abstract
  syntax tree, which can then be differentiated and inserted into a
  time stepping loop.}
\end{figure}

@@

\hypertarget{header-n3294}{%
\section{Representation Learning for
Reparameterization}\label{header-n3294}}

The central idea of this work is to obtain two new variables solve for
those directly which have no physical meaning other than simply being a
parameterization of the equation of state constraint. We want to select a
\(q_1\) and \(q_2\) to require no auxilary phase index to define the system
and no additional logic in the code. The sharp kinks and possible
discontinuities can be a part of the functions
\(\rho(q_1, q_2), p(q_1, q_2), h(q_1, q_2)\) and \(T(q_1, q_2)\).

@@

Instead of the painstaking work of parameterizing phase boundaries and
determining curve fits by hand, we rephrase the problem of representing
the equation of state by learning and autoencoder. We have an encoding
phase, \(E(\rho,p,e,T; a)\) and a decoding phase, \(D(u,v; b)\) that
forms an identity function with a compressed subspace:
\begin{equation}
\left\{ \begin{array}{c}
T\\ p\\ \rho\\ h
\end{array}\right\} \rightarrow  E \rightarrow 
\left\{ \begin{array}{c} q_1\\q_2 \end{array} \right\}\rightarrow D \rightarrow 
\left\{ \begin{array}{c}
T\\ p\\ \rho\\ h
\end{array}\right\}
\end{equation}
The autoencoder is solved for by optimizing its parameters using the
goal

@@

\[L(x)=\left\|x-D(E(x))\right\|^2_2+\lambda\left\|\frac{\partial E}{\partial x}\right\|_F^2\]

@@

The dataset is split into a standard $2/3-1/6-1/6$ division into
training-testing-validation groups. We note that we can always
manufacture more datapoints. We train by alternating phases. The first
step is minibatched stochastic descent on the entire deep network, assembling the losses on minibatches of the data, $X_{mini}$,
\begin{equation}
  L(X_{mini}) = \sum_{x\in X_{mini}} L(x)
  \end{equation}
to perform the out-of-the-box ADAM optimizer on all of the
parameters in the networks. (The box is
TensorFlow.) After a few epochs of this,
the training algorithm alternates to a few steps step of linear regression on the last layer of the network,
denoted by \(W^{dec}\), which directly controls the magnitudes of the values.
(The bias $b^{dec}$ is included in $W^{dec}$.) The linear
system that needs to be solved is easily generated by assembling the
gradient and Hessian of the total loss function,
\begin{align}
 \mathbf{G}\left(X_{train}\right) &= \sum_{x\in X_{train}}\frac{\partial L}{\partial W^{dec}} \\
\mathbf{H}\left(X_{train}\right) &= \sum_{x\in X_{train}}\frac{\partial^2 L}{\partial (W^{dec})^2}
\end{align}
and solving the linear system for an update to the last parameters.

\[[\mathbf{H}] \Delta W^{dec} = -\{\mathbf{G}\}\]

To form the hessian and gradient for this step, we sum over all of the
training data. (The contractive penalty term drops out of the this step because it does
not depend on \(W^{dec}\).)

\begin{equation}
\bar{H}_{ij} = H_{ij} + \left(\delta_{ij}\,\text{if}\,H_{ik}=0\forall k\right)
  \end{equation}

For this step in which only the back-end regression is performmed, each set of coefficients for the curves for each variable is independent from those of the other variables: e.g., the coefficients to the sets of polynomials that parameterize $p(q)$ do not depend on those for $T(q)$. This means that this step can be performmed in blocks, performing the linear step for each component independently, at a great reduction of cost. For the full system, 3/4 of the entries are zero. We did not bother performing this optimization, but note that the savings are quadratic as the number of coefficients grow for more complex equations of state. (This is not true for training the entire deep network: it is only true for this linear regression step on the sub-network.)

@@

Examining the $T,p,\rho$ surface allows for human interpretability, but
the decoder actually yields four new functions with two unknowns:
$D(q)=\{T(q),p(q),\rho(q),h(q)\}$. These surfaces can be plotted as
independent 2D functions: the surface of \ref{fig:classified} is
unfolded into the new parameterization in Figure \ref{fig:foursurfaces}.
\begin{figure}
  \centering
  \includegraphics{../slides/water_four_surfaces_plot.png}
  \caption{\label{fig:foursurfaces}The four 2D surfaces that make up
    the new parameterization. The learned surface captures the sharp
    kinks and vertical fronts in Figure \ref{fig:classifyingresult},
    but is an analytic and smooth function of the two learned unknowns.}
  \end{figure}

With the trained model, we solve for \(q_1(t)\) and \(q_2(t)\) such that:
\begin{align}
\partial_t \rho(q_1,q_2) & = \mathbf{k} \left( p_\infty-p(q_1,q_2) \right)  + r \\
\partial_t \left(\rho h(q_1,q_2)-p\right) &
                                            =\mathbf{k}_T(T_\infty-T(q_1,q_2))  + s + \frac{\rho h}{h}\left(\mathbf{k}_p(p_\infty - p) + r\right)
\end{align}
where the expressions for  \(\rho(q_1,q_2)\) etc. are the back of the
trained autoencoder,. The trained
autoencoder automatically removes the local constraint, satisfying
\begin{equation}
eos(D(q_1,q_2)) = 0 \, \forall \,q_1,q_2
\end{equation}
Thus, the system will always satisfy the constraints as $q$ is
evolved, even during intermediate iterative solution steps. Note again
that the balance laws are directly incorporated into the simulation
after loading the trained model.

To develop the rest of the discretization, we introduce the compact form:
\begin{align}
\partial_t m(T,p,\rho,h) &= r(T,p,\rho,h)\\
eos(T,p,\rho,h) &= 0
\end{align}
where $m=\{\rho,\rho h -p\}$ and $r$ is the right hand side, where
$eos(T,p,\rho,h)=0$ is the material constraint.

The decoder part of the network is inserted into the balance law
\begin{equation}
\partial_t (D(q)) = r(D(q))
\end{equation}
We can parameterize this equation with any time stepping scheme, .e.g. a
Runge-Kutta scheme. For subsurface flow, backward Euler is popular due
to its simplicity and L-stability, which we follow since the topic of
this paper is not exploring new time stepping schemes. This yields the
discretized algebraic equation for \(q_i\) and \(t+\Delta t\) given the
value \(q^0\) at \(t\),

\begin{equation}
0 = R(q^i) = m(q^{i}) - m(q^{0}) - \Delta t \, r(q^i)
\end{equation}
where we have called the equation \(R(q)\). By design \(R(q)\) is smooth
enough to solve with Newton's method without any statemachine logic:
\begin{equation}
\frac{\partial R}{\partial q} \Delta q^{k+1} = - R(q^k)
\end{equation}
with the update
\begin{equation}
q^{k+1} = q^k + \alpha\Delta q^{k+1}
\end{equation}
where $\alpha$ is an under relaxation parameter.

This equation is deceptively simple: the function $R(q)$ has complex logic encoded in it; if-then
statements appear if the model used activations such as rectifiers.
However, the complex material behavior which traditionally requires
the vast majority of programming was implemented automatically by the
models; implementing equations XX, XX, and XX requires only a few lines
of human-written code.

@@

The current implementation LatentSim does not support inputting via
saturations on the equilibria, a commonly used primary variable, but
this problem could be solved with future engineering work for a
non-experimental simulator.

For LatentSim, the input system is not yet as robust as a
fully-fledged user-facing simulator, but the inputs are specified in
the $T p \rho h$ space instead of the $q$ space.
The user (or test problem) specifies any two of the four values in a
dictionary format,
\begin{verbatim}
    p=9e5 T=20
    rho=500 T=20
\end{verbatim}
Then, the simulator must determine $q(t=0)$ from these two
values, but the econder only works when all four values are on the
surface. Let $s=\{T,p,\rho,\rho h\}$ be a four dimensional
variable of the stacked physical quantities. It is necessary to
actually solve the following nonlinear problem
on the decoder for the indices $i$ of the two specified values:
\begin{equation}
  D(q^0)_i = s^0_i \quad \text{for specified}\; i
\end{equation}
This equation has two rows indexing into the four state variables;
e.g. only the rows associated with $p$ and $T$ are solved if those are
specified, which is usually the case. The resulting system of
equations is solved with Newton's method. The two corresponding rows are cut out of
the tangent matrix to $D$,
\begin{equation}
K_{ij} = \frac{\partial D_i}{\partial q_j}  \quad \text{for
    specified}\; i, \text{for}\;j=1,2
\end{equation}
and the $2\times2$ linear system is solved for an update to $q$,
\begin{equation}
K_{ij}\Delta q_j = s^0_i-D(q)_i
\end{equation}
The iteration on $q$ is repeated until convergence, with under-relaxation used again to update:
 \begin{equation}
q^0_{k+1} =q^0_{k} + \alpha \Delta q_{k}
\end{equation}
with the under-relaxation parameter that is usually set to $\alpha \approx
0.1$.

>>>>

\jnlcitation{} %\cname{Queiruga, A.}\ctitle{Solution of DAEs on Latent Spaces},\cjournal{I. J. Num. Meth. Eng.}, \cvol{2019;00:1--2}

\keywords{}

@@

\begin{figure}
  \centering
  \includegraphics[width=3in]{../figures/cstr_vector.pdf}
  \caption{ \label{fig:cstr} Illustration of the Continuous
    Stirred-Tank    
  Reactor. The material in the reactor is exposed to a reservoir with
  pressure
  $p_\infty$ and temperature $T_\infty$ via a conduit with permeability
  and thermal conductivities $k_p$ and $k_T$, direct forced mass
  injection or withdrawal $r$ via a pump, or heating and cooling $s$
  via a heater/heat-exchanger. Stirring and insulation keeps the
  material at a uniform pressure and temperature. Gravity is
  neglected (though is usually the motivation for putting a feeding inlet at
the top and product outlet at the bottom.)}
\end{figure}

@@

Note that the definition of enthalpy as a per-mass quantity means the mixture needs to be weighted by the
density: $h_{mix}=\sum{\rho_a h_aS_a}/\sum{rho_aS_a}$.

But, what is a phase boundary? The traditional explanation is shown in
Figure \ref{fig:phaseboundaries}. The use of intensive properties breaks down:\footnote{https://en.wikipedia.org/wiki/Phase\_transition}
\begin{quote}
Phase transitions occur when the thermodynamic free energy of a system
is non-analytic for some choice of thermodynamic variables (cf. phases).
\end{quote}
{\bf Remark:} This interpretation breaks down with the existence of super-saturated phases in non-stable regions: e.g., it is possible to
heat water beyond it's boiling point in a microwave. We are concerned
with environments where this would not happen (the very rough and
heterogenous pores of rocks), so we have not thought of how to
integrate this behavior.

@@

This work focuses on the EOS of pure water. The most accurate
formulations were developed by the International Association for the
Properties of Water Steam (IAPWS) using a variety of experimental
sources for multiple phases \cite{wagner_iapws_2000,
  water_revised_2007, water_revised_2009}. We utilize the 1997
formulation for water gas, liquid, and supercritical phases, and the
2006 formulation for the Ice Ih phase to build the dataset.\footnote{The
implementation is publicly available as a stand-alone library at \url{https://github.com/afqueiruga/equations\_of\_state}.} (Their
painstaking work in determining the forms of decent fits is a great
motivation for a 21st century approach.)
We can also assemble the surface of possible $T,p,\rho$ coordinates,
shown in Figure \ref{fig:pTrho}.
\begin{figure}
\centering
\includegraphics[width=4in]{../figures/phase_diagram.png}
\caption{\label{fig:phaseboundaries}Phase boundaries for water on the $T-p$ diagram. The standard
interpretation is that the properties of the material change
drastically when crossing this boundary. Note how it is possible to form a path from a
definitively gas state to a definitively liquid state without crossing
a phase boundary. The supercritical regime has no clear boundaries,
but requires a separate curve fit in practice due to numerical challenges.}
\end{figure}

@@

{\bf Remark:} Many popular forms of EOS do not even have closed form
expressions. The family of cubic equations of state for non-ideal
gases, such as the Peng-Robinson equation \cite{peng_new_1976}, are
expressed as functions for $p$ of $V$ (a proxy for $\rho$) and
$T$. Since $\rho$ is rarely a primary variable, these expressions must be inverted for
$\rho$ given $p$ by solving the cubic equation. This can be performed
analytically, as is done in TOUGH+, but often polynomial roots can
be best solved numerically. When the equations become more
complicated, such as when there are gas mixtures, the equations of
state must be solved numerically. TOUGH+ employs a Richardson
iteration \cite{moridis_users_2014-2}.
Note how this is a numerical
iteration {\em inside} of another numerical iteration that is trying
to differentiation this equation state. Approximations
abound--the mathematical theory is not straight forward. Passing
automatic differentiation objects through these computations is
difficult in practice. Equations are
often solved faster and more accurately by numerical
approximations--why not completely rewrite the equations?

@@

The typical methodology is to use the empirical relations for density
and enthalpy \(h\) as a function of pressure and temperature and solve
the DAE for \(p\) and \(T\) implicitly:
\begin{align}
\text{Solve for}\, p(t)\, \text{and}\, T(t)\, \text{such that:}\\
\partial_t \rho(p,T) & = \nabla \cdot \mathbf{k}\nabla p + r\\
\partial_t \rho(p,T) h(p,T)-p & = \nabla \cdot \mathbf{k'}\nabla T + s
\end{align}
The complication is that the functions \(\rho(p,T)\) and \(u(p,T)\)
are not well defined functions due to the presence of phase changes that
yield sharp discontinuities in \(p,T\), as shown in Figure 1 for
water.
Each state requires a potentially different set of primary
variables. For a pure water system, this would have to be:
\begin{center}
\begin{tabular}{l|c|c}
  State & $X_1$ & $X_2$ \\
  \hline
  Ice & $p$ & $T$ \\
  Liquid & $p$ & $T$ \\
  Gas & $p$ & $T$ \\
  Supercritical & $\rho$ & $T$ \\
  Ice-Liquid & $S_{ice}$ & $T$ \\
  Ice-Gas & $S_{ice}$ & $T$ \\
  Gas-Liquid & $S_{gas}$ & $T$ \\
  Ice-Gas-Liquid & $S_{ice}$ & $S_{gas}$ 
\end{tabular}
\end{center}
Solving the balance laws with this caveat requires the incorporation
of a statemachine. Representing the material with the known
representations thus requires two variables and an additional state
tag. For the complete states of water, the states and their possible
transitions is shown in Figure \ref{fig:statemachine}.
  \begin{figure}
    \centering
    \includegraphics[width=4in]{../slides/phase_4.png}
    \caption{\label{fig:statemachine}States and possible transitions for water.}
\end{figure}
The discretized balance equations become a nonlinear function of
continuous variables $X$ and discrete indices $phase$:
\begin{equation}
R(X ; phase) = 0
\end{equation}
where the meaning of $X$ is dictated by $phase$. Solving this equation
requires first iterating on the differentiable part by solving the
Jacobian matrix,
\begin{equation}
  \frac{\partial R}{\partial X}\Delta X^{k+1/2} = -R(X^k ; phase^k)
\end{equation}
and the iterating the non-differentiable state machine part:
\begin{equation} \label{eq:fsm}
  phase^{k+1},X^{k+1} = statemachine(X^{k+1/2} ; phase^k)
\end{equation}
The phase switching logic has the form of a switch case illustrated in
Algorithm \ref{alg:statemachine}. This converges quadratically in
situations where no phases changes occur, but execution of
Eq. \ref{eq:fsm} breakdown into linear convergence or even divergence.
This often contains the presence of ``tricks'' such as overshoots to
push the grid blocks over the edge, $P_{new}=(1+10^{-6} )P_{old}$, or
artificially smoothing the phase boundaries
\citep{moridis_simulation_2019}. 
 Besides the computational difficulties, from a software
engineering standpoint, the long and winding logical branches are very
difficult to maintain and extend with new physics. The presence of the state machine and implicit equations inside of the
material descriptions also requires numerical differentiation, which is
a further numerical challenge. The engineering
difficulty further makes it difficult to improve the efficiencies of the implementations.

@@

An archetypical constraint problem is the pendulum. Posed as dynamics
problem with a Lagrange-multipler-esque centripetal force, it
has three unknowns with two second-order equations and a constraint,
\begin{align}
\text{Solve for}\, x(t), \, y(t), \, f(t) \, \text{satisfying:} \nonumber \\
m \ddot{x} & = f x/\sqrt{x^2+y^2} \\
m \ddot{y} & = f y/\sqrt{x^2+y^2} - m g \\
  x^2 + y^2 & = R^2
  \text{given } x,y, \dot{x},\dot{y} \text{ at } t=0. \nonumber
\end{align}
Note that we are considering the nonlinear pendulum.

@@

\begin{figure}
  \center
\includegraphics[width=5in]{../figures/autoencoder_balance_detailed.png}
  \caption{\label{fig:balance}Illustration of the insert of the encoder to form the
  description of the system of equation. The equations
  of motion are generated in a declarative programming methodology,
  where the trained decoder can be extracted and inserted into the abstract
  syntax tree, which can then be differentiated and inserted into a
  time stepping loop.}
\end{figure}

@@

\hypertarget{header-n3294}{%
\section{Representation Learning for
Re-parameterization}\label{header-n3294}}

The central idea of this work is to obtain two new variables solve for
those directly which have no physical meaning other than simply being a
parameterization of the equation of state constraint. We want to select a
\(q_1\) and \(q_2\) to require no auxiliary phase index to define the system
and no additional logic in the code. The sharp kinks and possible
discontinuities can be a part of the functions
\(\rho(q_1, q_2), p(q_1, q_2), h(q_1, q_2)\) and \(T(q_1, q_2)\).

@@

Instead of the painstaking work of parameterizing phase boundaries and
determining curve fits by hand, we rephrase the problem of representing
the equation of state by learning and autoencoder. We have an encoding
phase, $E(\rho,p,e,T; a)$ and a decoding phase, $D(u,v; b)$ that
forms an identity function with a compressed subspace:
\begin{equation}
\left\{ \begin{array}{c}
T\\ p\\ \rho\\ h
\end{array}\right\} \rightarrow  E \rightarrow 
\left\{ \begin{array}{c} q_1\\q_2 \end{array} \right\}\rightarrow D \rightarrow 
\left\{ \begin{array}{c}
T\\ p\\ \rho\\ h
\end{array}\right\}
\end{equation}
The autoencoder is solved for by optimizing its parameters using the
goal

@@

\begin{equation}
L(x)=\left\|x-D(E(x))\right\|^2_2+\lambda\left\|\frac{\partial E}{\partial x}\right\|_F^2
\end{equation}

@@

The dataset is split into a standard $2/3-1/6-1/6$ division into
training-testing-validation groups. We note that we can always
manufacture more datapoints. We train by alternating phases. The first
step is mini-batched stochastic gradient descent on the entire deep network, assembling the losses on minibatches of the data, $X_{mini}$,
\begin{equation}
L(X_{mini}) = \sum_{x\in X_{mini}} L(x)
\end{equation}
to perform the out-of-the-box ADAM optimizer on all of the
parameters in the networks. (The box is
TensorFlow.) After a few epochs of this,
the training algorithm alternates to a few steps step of linear regression on the last layer of the network,
denoted by \(W^{dec}\), which directly controls the magnitudes of the values.
(The bias $b^{dec}$ is included in $W^{dec}$.) The linear
system that needs to be solved is easily generated by assembling the
gradient and Hessian of the total loss function,
\begin{align}
\mathbf{G}\left(X_{train}\right) &= \sum_{x\in X_{train}}\frac{\partial L}{\partial W^{dec}} \\
\mathbf{H}\left(X_{train}\right) &= \sum_{x\in X_{train}}\frac{\partial^2 L}{\partial (W^{dec})^2}
\end{align}
and solving the linear system for an update to the last parameters.

\begin{equation}
[\mathbf{H}] \Delta W^{dec} = -\{\mathbf{G}\}\]
\end{equation}

To form the hessian and gradient for this step, we sum over all of the
training data. (The contractive penalty term drops out of the this step because it does
not depend on \(W^{dec}\).)
\begin{equation}
\bar{H}_{ij} = H_{ij} + \left(\delta_{ij}\,\text{if}\,H_{ik}=0\forall k\right)
\end{equation}
For this step in which only the back-end regression is performed, each set of coefficients for the curves for each variable is independent from those of the other variables: e.g., the coefficients to the sets of polynomials that parameterize $p(q)$ do not depend on those for $T(q)$. This means that this step can be performed in blocks, performing the linear step for each component independently, at a great reduction of cost. For the full system, 3/4 of the entries are zero. We did not bother performing this optimization, but note that the savings are quadratic as the number of coefficients grow for more complex equations of state. (This is not true for training the entire deep network: it is only true for this linear regression step on the sub-network.)

@@

\subsection{Single Grid Block}

\begin{figure}
  \centering
  \includegraphics{../slides/water_four_surfaces_plot.png}
  \caption{\label{fig:foursurfaces}The four 2D surfaces that make up
    the new parameterization. The learned surface captures the sharp
    kinks and vertical fronts in Figure \ref{fig:classifyingresult},
    but is an analytic and smooth function of the two learned unknowns.}
  \end{figure}

Examining the $T,p,\rho$ surface allows for human interpretability, but
the decoder actually yields four new functions with two unknowns:
$D(q)=\{T(q),p(q),\rho(q),h(q)\}$. These surfaces can be plotted as
independent 2D functions: the surface of \ref{fig:classified} is
unfolded into the new parameterization in Figure
\ref{fig:foursurfaces}. These surfaces were designed to be smooth
enough to solve effectively.
With the trained model, we solve for \(q_1(t)\) and \(q_2(t)\) such that:
\begin{align}
\partial_t \rho(q_1,q_2) & = \mathbf{k} \left( p_\infty-p(q_1,q_2) \right)  + r \label{eq:massbalanceq}\\
\partial_t \left(\rho h(q_1,q_2)-p\right) &
                                            =\mathbf{k}_T(T_\infty-T(q_1,q_2))+ s + \frac{\rho h}{h}\left(\mathbf{k}_p(p_\infty - p) + r\right) \label{eq:heatbalanceq}
\end{align}
where the expressions for  \(\rho(q_1,q_2)\) etc. are the back of the
trained autoencoder,. The trained
autoencoder automatically removes the local constraint, satisfying
\begin{equation}
eos(D(q_1,q_2)) = 0 \, \forall \,q_1,q_2
\end{equation}
Thus, the system will always satisfy the constraints as $q$ is
evolved, even during intermediate iterative solution steps. Note again
that the balance laws are directly incorporated into the simulation
after loading the trained model.

To develop the rest of the discretization, we introduce the compact form:
\begin{align}
\partial_t m(T,p,\rho,h) &= r(T,p,\rho,h)\\
eos(T,p,\rho,h) &= 0
\end{align}
where $m=\{\rho,\rho h -p\}$ and $r$ is the right hand side, where
$eos(T,p,\rho,h)=0$ is the material constraint. The decoder part of the network is inserted into the balance law
\begin{equation}
\partial_t (D(q)) = r(D(q))
\end{equation}
We can parameterize this equation with any time stepping scheme, e.g. a
Runge-Kutta scheme. For subsurface flow, backward Euler is popular due
to its simplicity and L-stability. We make the same choice because the topic of
this paper is not exploring new time stepping schemes. This yields the
discretized algebraic equation for \(q_i\) and \(t+\Delta t\) given the
value \(q^0\) at \(t\),
\begin{equation}
0 = R(q^i) = m(q^{i}) - m(q^{0}) - \Delta t \, r(q^i) \label{eq:bweuler}
\end{equation}
where we have called the equation \(R(q)\). By design \(R(q)\) is smooth
enough to solve with Newton's method without any statemachine logic:
\begin{equation}
\frac{\partial R}{\partial q} \Delta q^{k+1} = - R(q^k) \label{eq:jacobianq}
\end{equation}
with the update
\begin{equation}
q^{k+1} = q^k + \alpha\Delta q^{k+1}
\end{equation}
where $\alpha$ is an under relaxation parameter.

This equation is deceptively simple: the function $R(q)$ has complex logic encoded in it; if-then
statements appear if the model used activations such as rectifiers.
However, the complex material behavior which traditionally requires
the vast majority of programming was implemented automatically by the
models; implementing equations \ref{eq:massbalanceq}, \ref{eq:heatbalanceq}, and \ref{eq:bweuler } requires only a few lines
of human-written code. The iteration for \ref{eq:jacobianq} can be constructed entirely with automatic differentiation because the model is fully differentiable.

@@

For LatentSim, the input system is not yet as robust as a
fully-fledged user-facing simulator, but the inputs are specified in
the $T p \rho h$ space instead of the $q$ space.
The user (or test problem) specifies any two of the four values in a
dictionary format,
\begin{verbatim}
    p=9e5 T=20
    rho=500 T=20
\end{verbatim}
The current implementation LatentSim does not support inputting via
saturations on the equilibria, a commonly used primary variable, but
this problem could be solved with future engineering work for a
non-prototype simulator.
Then, the simulator must determine $q(t=0)$ from these two
values, but the encoder only works when all four values are on the
surface. Let $s=\{T,p,\rho,\rho h\}$ be a four dimensional
variable of the stacked physical quantities. It is necessary to
actually solve the following nonlinear problem
on the decoder for the indices $i$ of the two specified values:
\begin{equation}
  D(q^0)_i = s^0_i \quad \text{for specified}\; i
\end{equation}
This equation has two rows indexing into the four state variables;
e.g. only the rows associated with $p$ and $T$ are solved if those are
specified, which is usually the case. The resulting system of
equations is solved with Newton's method. The two corresponding rows are cut out of
the tangent matrix to $D$,
\begin{equation}
K_{ij} = \frac{\partial D_i}{\partial q_j}  \quad \text{for
    specified}\; i, \text{for}\;j=1,2
\end{equation}
and the $2\times2$ linear system is solved for an update to $q$,
\begin{equation}
K_{ij}\Delta q_j = s^0_i-D(q)_i
\end{equation}
The iteration on $q$ is repeated until convergence, with under-relaxation used again to update:
 \begin{equation}
q^0_{k+1} =q^0_{k} + \alpha \Delta q_{k}
\end{equation}
with the under-relaxation parameter that is usually set to $\alpha \approx
0.1$.

The initial condition problem is not guaranteed to converge, however,
and often does not.
For example, if the initial guess was in the liquid phase and the target point
in the EOS space is in a solid, the iteration will diverge.
To address this issue, the solver performs multiple starts with an initial condition in each
phase and equilibrium, discard points that diverge, and then picks the
\(q\) that is closest to the two specified components. The multiple
start points are specified at EOS development time to form a set of
values that represent middle points for each phase, \footnote{The multistart method could be specified without any prior knowledge
with random points, but picking start points with prior knowledge was necessary to test the cause of the divergence.}
\begin{equation}
\left\{s_{guess}\right\} = \left\{ \left(\begin{array}{c}
\bar{T}_{gas}\\
\bar{p}_{gas}\\
\bar{\rho}_{gas}\\
\bar{\rho h}_{gas}
\end{array}\right),\left(\begin{array}{c}
\bar{T}_{liq}\\
\bar{p}_{liq}\\
\bar{\rho}_{liq}\\
\bar{\rho h}_{liq}
\end{array}\right),\left(\begin{array}{c}
\bar{T}_{ice}\\
\bar{p}_{ice}\\
\bar{\rho}_{ice}\\
\bar{\rho h}_{ice}
\end{array}\right)... \right\}
  \end{equation}
Each of the possible guess points is used to fill in the two
unspecified values as guesses and then passed into the encoder.
E.g. if $p$ and $T$ were specified, each possible $\bar{\rho}$ and
$\bar{h}$ combination from the above sets are filled in to provide
multiple guesses for $q$,
\begin{equation}
s^0 = \left(\begin{array}{c}
T^*(0)\\
p^*(0)\\
\bar{\rho}_{guess}\\
\bar{h}_{guess}
\end{array}\right);\quad q^0_{guess}=E\left(s^0\right)
 \end{equation}
These sets of $q^0_{guess}$ are then iterated upon using the Newton's
 method procedure.
